# Model 2 - Mixed Effects Modelling

Main Idea: For model 1, we took every observation as i.i.d. However, there may be other effects that we haven't measured in our model such as team effects or even player effects. Our data consists of Squad Size and Average Market Value of each team both of which are Normalized by Average. There is 1 data point for each team for each year. This means that there are team effects and possibly year effects that we can encounter. The team effect is the first and main focus since some teams may have special patterns and trends. For example, if a team finishes 18, 19, or 20th in the league they are relegated to the EFL Champsionship (the 2nd division of English Football). This means that there are only a handful of teams that have remained in the Premier League since its inception (1992-93). This team effect may be crucial!

Resources/Links for Inspiration and Learning

-   <https://meghan.rbind.io/blog/2022-06-28-a-beginners-guide-to-mixed-effects-models/>

-   <https://mfviz.com/hierarchical-models/>

```{r}
# Install packages if needed
# install.packages(c("lme4", "lmerTest", "dplyr", "ggplot2"))

library(lme4)      # For mixed effects models
library(lmerTest)  # For p-values
library(dplyr)     # For data manipulation
library(ggplot2)   # For visualization
```

1.  Load and Prepare the Data

```{r}
# Load data
df <- read.csv("/Users/apendela10/STAT482/statistics-capstone/r-modelling/transfermarkt_financial_final.csv")

# Check data structure
cat("Total observations:", nrow(df), "\n")
cat("Number of unique teams:", length(unique(df$Team)), "\n")

# Check team representation
team_counts <- df %>%
  group_by(Team) %>%
  summarise(n_obs = n()) %>%
  arrange(desc(n_obs))

cat("\nMin observations per team:", min(team_counts$n_obs), "\n")
cat("Max observations per team:", max(team_counts$n_obs), "\n")
cat("Mean observations per team:", round(mean(team_counts$n_obs), 1), "\n")

# Count teams by observation frequency
cat("\nTeam representation:\n")
cat("  Teams with 1 observation: ", sum(team_counts$n_obs == 1), "\n")
cat("  Teams with 2 observations: ", sum(team_counts$n_obs == 2), "\n")
cat("  Teams with 3-5 observations:", sum(team_counts$n_obs >= 3 & team_counts$n_obs <= 5), "\n")
cat("  Teams with 6+ observations:", sum(team_counts$n_obs >= 6), "\n")
```

This shows that we have 21 years of data which end in the 2024-25 season. With 3 teams relegated each year and 3 teams being promoted from the championship, there are 44 unique teams in the Premier League in this 21 year time span.

2.  Fitting the Mixed Effects Model

```{r}
# Fit model with random intercept for Team
model <- lmer(Points ~ Squad_Size_Normalized_by_Avg + 
                Average_Market_Value_Normalized_by_Avg + 
                (1 | Team),
              data = df,
              REML = TRUE)

# Display summary
cat("\nModel Summary:\n")
print(summary(model))
```

3.  Extracting Model Parameters

```{r}
# Fixed effects
fixed_effects <- fixef(model)
print(fixed_effects)

# Random effects (team-specific intercepts)
random_effects <- ranef(model)$Team
random_effects$Team <- rownames(random_effects)
colnames(random_effects)[1] <- "Random_Intercept"
random_effects <- random_effects[order(-random_effects$Random_Intercept), ]

# Calculate ICC
variance_components <- as.data.frame(VarCorr(model))
between_var <- variance_components$vcov[1]  # Team variance
within_var <- variance_components$vcov[2]   # Residual variance
icc <- between_var / (between_var + within_var)

cat("Between-team variance (τ²):", round(between_var, 3), "\n")
cat("Within-team variance (σ²):  ", round(within_var, 3), "\n")
cat("ICC (Intraclass Correlation):", round(icc, 3), "\n")
cat(sprintf("\n→ %.1f%% of variance is BETWEEN teams\n", icc * 100))
cat(sprintf("→ %.1f%% of variance is WITHIN teams (year-to-year)\n", (1-icc) * 100))
```

4.  Visualizing the Random Effects

```{r}

random_effects$color <- ifelse(random_effects$Random_Intercept > 0, "positive", "negative")

p <- ggplot(random_effects, aes(x = reorder(Team, Random_Intercept), 
                                y = Random_Intercept, 
                                fill = color)) +
  geom_bar(stat = "identity", alpha = 0.7) +
  coord_flip() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black", size = 1) +
  scale_fill_manual(values = c("negative" = "red", "positive" = "green")) +
  labs(title = "Team-Specific Random Effects",
       subtitle = "Deviation from Average After Controlling for Predictors",
       x = "Team",
       y = "Random Intercept (Team Effect)") +
  theme_minimal() +
  theme(legend.position = "none",
        plot.title = element_text(face = "bold", size = 14),
        axis.text.y = element_text(size = 8))

print(p)

# Save plot
ggsave("team_random_effects.png", plot = p, width = 10, height = 12, dpi = 300)
cat("✓ Plot saved as 'team_random_effects.png'\n")

```

5.  Loading 2025-26 Season Data

```{r}
# Load 2025 inference data
df_2025 <- read.csv("/Users/apendela10/STAT482/statistics-capstone/r-modelling/2025_inference.csv")

cat("Number of teams in 2025-26 season:", nrow(df_2025), "\n")
cat("\n2025-26 Teams:\n")
print(df_2025$Team)

predict_with_uncertainty <- function(new_data, model, random_effects) {
  # Get fixed effects
  fixed_pred <- predict(model, newdata = new_data, re.form = NA)  # Population level
  
  # Get variance components
  vc <- as.data.frame(VarCorr(model))
  tau_sq <- vc$vcov[1]  # Between-team variance
  sigma_sq <- vc$vcov[2]  # Within-team variance
  
  # Prepare results
  results <- data.frame(
    Team = new_data$Team,
    Squad_Size_Norm = new_data$Squad_Size_Normalized_by_Avg,
    Market_Value_Norm = new_data$Average_Market_Value_Normalized_by_Avg,
    Fixed_Effect_Pred = fixed_pred
  )
  
  # Check if team has random effect
  results$Has_Random_Effect <- tolower(results$Team) %in% tolower(rownames(random_effects))
  
  # Add team-specific predictions
  results$Team_Effect <- 0
  results$Predicted_Points <- results$Fixed_Effect_Pred
  results$SE <- sqrt(sigma_sq + tau_sq)  # Default to new team SE
  results$Uncertainty <- "Higher (New Team)"
  
  for (i in 1:nrow(results)) {
    team_lower <- tolower(results$Team[i])
    if (team_lower %in% tolower(rownames(random_effects))) {
      # Team has random effect
      team_effect <- random_effects[tolower(rownames(random_effects)) == team_lower, "Random_Intercept"]
      results$Team_Effect[i] <- team_effect
      results$Predicted_Points[i] <- results$Fixed_Effect_Pred[i] + team_effect
      results$SE[i] <- sqrt(sigma_sq)  # Known team has lower SE
      results$Uncertainty[i] <- "Lower (Known Team)"
    }
  }

  # Add 95% prediction intervals
  results$CI_Lower <- results$Predicted_Points - 1.96 * results$SE
  results$CI_Upper <- results$Predicted_Points + 1.96 * results$SE
  
  return(results)
}
```

```{r}
predictions <- predict_with_uncertainty(df_2025, model, random_effects)

# Sort by predicted points
predictions <- predictions %>%
  arrange(desc(Predicted_Points)) %>%
  mutate(Predicted_Position = row_number())

predictions_display <- predictions %>%
  select(Predicted_Position, Team, Predicted_Points, Team_Effect, Uncertainty)

print(predictions_display, row.names = FALSE)
```

```{r}
# Prediction Intervals

predictions_ci <- predictions %>%
  select(Predicted_Position, Team, Predicted_Points, CI_Lower, CI_Upper, Uncertainty)

print(predictions_ci, row.names = FALSE)
```

```{r}
# Check for convergence warnings
if (!is.null(model@optinfo$conv$lme4$messages)) {
  cat("⚠ Convergence warnings detected:\n")
  print(model@optinfo$conv$lme4$messages)
} else {
  cat("✓ Model converged successfully with no warnings\n")
}

# AIC and BIC
cat("\nModel Fit Statistics:\n")
cat("AIC:", round(AIC(model), 2), "\n")
cat("BIC:", round(BIC(model), 2), "\n")
cat("Log-Likelihood:", round(logLik(model), 2), "\n")
```

Ideas to improve the model

-   year effect...

Ideas to improve this model with further data...

-   attendance data (/league_table_1992-1993_to_2025-2026.csv)

<!-- -->

-   on field performance data from the season before

    -   top scorer (/league_table_1992-1993_to_2025-2026.csv)

    -   xG,xGA,xGD (/league_table_1992-1993_to_2025-2026.csv)

    -   
